---
title: "Week 1 Assignment"
author: "John Navarro"
date: "September 28, 2017"
output: pdf_document
---

# Study on Smokers

This project helps understanding conditional distributions.
The following data set is in library(faraway).
In it participating women are categorized into groups of smokers or not, 7 age groups and groups of dead or alive after 20 years of study.
```{r}
# install packages and download dataset
library(faraway)
data(femsmoke)
dim(femsmoke)  #28 x 4
# print head of the data set
(femsmoke)
```
Create joint distribution of 3 counts variables:

v - "smoker yes"=1 or "smoker no"=0
u - "dead yes"=1 or "dead no"=0
w - age category
w=1: 18-24
w=2: 25-34
w=3: 35-44
w=4: 45-54
w=5: 55-64
w=6: 65-74
w=7: 75+
```{r}
# subset the data by dead category, extract the group counts
mat.dead.yes <- subset(femsmoke, dead == "yes")[,1]
mat.dead.no <- subset(femsmoke, dead == "no")[,1]
dim(mat.dead.yes) <- dim(mat.dead.no) <- c(2,7)
# combine the goup counts in to an array
femsmoke.joint <- array(rep(NA,28), dim = c(2,7,2))
femsmoke.joint[,,1] <- mat.dead.yes
femsmoke.joint[,,2] <- mat.dead.no
# name the variables of the array
dimnames(femsmoke.joint) <- list(c("smoke.yes","smoke.no"),c("age.1", "age.2" ,"age.3", "age.4", "age.5", "age.6", "age.7"), c("dead.yes","dead.no"))
# print the array of the group counts
femsmoke.joint
# Calculate the number of observations
N <- sum(femsmoke.joint)
# create the array of probabilities
femsmoke.joint.p <- femsmoke.joint/N
femsmoke.joint.p
# check that prob sum to 1
sum(femsmoke.joint.p)
```
Create marginal distribution for u (dead)
```{r}
(uMarginal.dead <- apply(femsmoke.joint.p,3,sum))
# check sums to 1
sum(uMarginal.dead)
```
Create marginal distribution for v(smoke)
```{r}
(vMarginal.smoke <- apply(femsmoke.joint.p,1,sum))
# check that sums to 1
sum(vMarginal.smoke)
```
Create marginal distribution for w (age)
```{r}
(wMarginal.age <- apply(femsmoke.joint.p,2,sum))
# chekc that sums to 1
sum(wMarginal.age)
```
Create conditional distribution p(w,v|u="alive")=p(smoke,age|alive)
```{r}
(cond.v.w.given.uAlive <- femsmoke.joint.p[,,"dead.no"]/uMarginal.dead["dead.no"])
# check that sums to 1
sum(cond.v.w.given.uAlive)
```
Create conditional distribution p(v|u="alive")=p(smoke|alive)
```{r}
(cond.v.given.uAlive <- apply(femsmoke.joint.p[,,"dead.no"],1,sum)/uMarginal.dead["dead.no"])
# check that sums to 1
sum(cond.v.given.uAlive)
```
Create conditional distribution p(w|u="alive",v="smoker")=p(age|alive,smoke)
```{r}
(cond.w.given.uAlive.vSmoke <- femsmoke.joint.p["smoke.yes",,"dead.no"]/cond.v.given.uAlive["smoke.yes"]/uMarginal.dead["dead.no"])
# check that sums to 1  
sum(cond.w.given.uAlive.vSmoke)  
```
Compare the vectors p(w|v2,u1)p(v2|u1)p(u1) and p(w,v,u)[,v2,u1]
```{r}
rbind(uMarginal.dead["dead.no"]*cond.v.given.uAlive["smoke.yes"]*cond.w.given.uAlive.vSmoke,
      femsmoke.joint.p["smoke.yes",,"dead.no"])
```


Let the marginal distribution for age group be p(w)p(w) estimated marginal distribution from the sample:
```{r}
wMarginal.age
```

Given simulated age group, simulate variable v using conditional distribution p(v|w),i.e. using probabilities p(smoke.yes|age), p(smoke.no|age).

Given simulated variables for age and for smoke, simulate mortality variable using distribution p(dead|v,w), p(alive|v,w).

Using the described procedure simulate outcomes for 100 participants.
Use seed set.seed(284) for comparison.

```{r}
# simulate number of observations in age groups 
set.seed(284)
sim.age <- sample(1:7,100, replace=T,prob=wMarginal.age)
# given simulated age group, simulate smoke variable
# calculate joint prob of smoke and age
joint.smoke.age <- apply(femsmoke.joint.p,c(1,2),sum)
# calculate the marginal matrix
marg <- rbind(colSums(joint.smoke.age), colSums(joint.smoke.age))
# calculate the conditional prob of smoke given age
(cond.smoke.given.age <- joint.smoke.age/marg)
# simulate the smoke variable given ages
#set.seed(284)
sim.smoke <- sapply(sim.age, function(z) sample(1:2,1, replace=T,
                                                prob=cond.smoke.given.age[,z]))
temp.vars <- cbind(sim.smoke,sim.age)
# Simulate mortality given age and smoke
#set.seed(284)
sim.dead <-apply(temp.vars,1,function(z) sample(1:2,1,replace=T, prob=femsmoke.joint.p[z[1],z[2],]))
# combine the data as simulatedData
simulatedData <- cbind(ages=sim.age, smokers=sim.smoke, dead = sim.dead-1)
```



```{r}
# return the head of the simulated data
head(simulatedData,25)
```

```{r}

```

