---
title: "BAWeek1Workshop1"
author: "John Navarro"
date: "September 27, 2017"
output: pdf_document
---
# 1. Joint distribution:3-dimensional Case

## 1.1 Creating 3-way array

Create a template for 3 dimensional distribution table

```{r}
# create vectors of variable names
u<-c("u1","u2")
v<-c("v1","v2","v3")
w<-c("w1","w2","w3")
# paste u0 to each string combination of v's and w's: vector of strings
matr.u0<-paste("u0",outer(v,w,paste,sep=","),sep=",")
# reshape into matrix
dim(matr.u0)<-c(3,3)
# paste u1 to each string combination of v's and w's: vector of strings
matr.u1<-paste("u1",outer(v,w,paste,sep=","),sep=",")
dim(matr.u1)<-c(3,3)
# reshape into matrix
matr.u0
```

```{r}
matr.u1
```

```{r}
# read in data
dataPath <- "C:/Users/JohntheGreat/Documents/MSCA/BayesianMethods/Week1_Probability"
data3way<-read.csv(file=paste(dataPath,"3WayData.csv",sep="/"))
dim(data3way) # 100 x 3
head(data3way)
```

```{r}
# use table to aggregate # of times each combination appears, when u=0
mat.u0<-table(subset(data3way,u==0)[,1],subset(data3way,u==0)[,2])
# use table to aggregate # of times each combination appears, when u=1
mat.u1<-table(subset(data3way,u==1)[,1],subset(data3way,u==1)[,2])
# return both tables
mat.u0
mat.u1
```

```{r}
idx.v1<-data3way$v==1
idx.w1<-data3way$w==1
idx.u1<-data3way$u==1
sum(idx.v1*idx.w1*idx.u1) #element (1,1) of mat.u1
```

```{r}
idx.v2<-data3way$v==2
sum(idx.v2*idx.w1*idx.u1) #element (1,2) of mat.u1
```

```{r}
idx.w2<-data3way$w==2
sum(idx.v1*idx.w2*idx.u1) #element (2,1) of mat.u1
```

```{r}
# assign column and row names
colnames(mat.u1)<-colnames(mat.u0)<-c("v1","v2","v3")
rownames(mat.u1)<-rownames(mat.u0)<-c("w1","w2","w3")
# create the array of NAs
data3way.array<-array(rep(NA,18),dim=c(3,3,2),dimnames=list(paste("w",1:3,sep=""),
                                                            paste("v",1:3,sep=""),
                                                            paste("u",0:1,sep="")))
# populate the NA array with totals from mat.u0 and mat.u1
data3way.array[,,1]<-mat.u0
data3way.array[,,2]<-mat.u1
# print the array
data3way.array
```
Create 3-dimensional joint distribution.
```{r}
# Sum all the occurances, should equal 100 observations
N<-sum(data3way.array)
# Create an array of probabilities
(data3way.array.p<-data3way.array/N)
```

## 1.2 Marginal Distributions

Create marginal distribution for u as vector uMarginal
```{r}
# marginal distribution, sum probabilities accross arrays
(uMarginal <- apply(data3way.array.p,3,sum))
```
Create marginal distribution for v as vector vMarginal.
```{r}
# sum probabilities across columns 
(vMarginal <- apply(data3way.array.p, 2, sum))
```
Create marginal distribution for w as vector wMarginal.
```{r}
# sum probabilities across rows
(wMarginal <- apply(data3way.array.p, 1, sum))
```

## 1.3 Conditional Distributions

Create conditional distribution p(w,v|u=1)as matrix cond.v.w.given.u1.
```{r}
# conditional disstribution = joint distribution/marginal distribution
# Joint dist of w and v , given u=1. divided by marginal dist of u1
cond.v.w.given.u1 <- data3way.array.p[,,"u1"]/uMarginal["u1"]
cond.v.w.given.u1
# should sum to 1
sum(cond.v.w.given.u1)
```
Create conditional distribution p(v|u=1) as vector cond.v.given.u1

```{r}
# cond distr of v's given u=1, divided by marginal dist of u=1
cond.v.given.u1 <- apply(data3way.array.p[,,"u1"], 2, sum)/uMarginal["u1"]
cond.v.given.u1
# should sum to 1
sum(cond.v.given.u1)
```

Create conditional distribution p(w|v=2,u=1) as vector cond.w.given.u1.v2.
```{r}
# cond distr of w given v=2 and u=1, divided by marginal distr of u=1
cond.w.given.u1.v2 <- data3way.array.p[,"v2","u1"]/cond.v.given.u1["v2"]/uMarginal["u1"]
cond.w.given.u1.v2
# should sum to 1
sum(cond.w.given.u1.v2)
```
Compare the vectors p(w|v2,u1)p(v2|u1)p(u1) and p(w,v,u)[,v2,u1]
```{r}
rbind(uMarginal["u1"]*cond.v.given.u1["v2"]*cond.w.given.u1.v2,data3way.array.p[,"v2","u1"])
```

In general, p(w,v,u)=p(w|v,u)p(v|u)p(u)

#2 Simulation Using Conditional Distributions

Let the marginal distribution for random variable u be Bernoulli with p(u=0)=0.55, p(u=1)=0.45
Let conditional distributions for random variables (v|u=0) and (v|u=1), taking values 1,2,3 be
```{r}
# conditional distribution of v given u=0
(pCond.v.given.u0<-c(.7,.2,.1))
# conditional distribution of v given u=1
(pCond.v.given.u1<-c(.1,.2,.7))
```
Let random variable (w|v,u) take values 1,2,3 with probabilities p(w|v,u), given by the following:
```{r}
p.given.u0.v1<-c(.3,.3,.4)
p.given.u0.v2<-c(.5,.3,.2)
p.given.u0.v3<-c(.6,.2,.2)
p.given.u1.v1<-c(.2,.3,.5)
p.given.u1.v2<-c(.2,.2,.6)
p.given.u1.v3<-c(.1,.7,.2)
```
Simulate joint sample (w,v,u) of lenth n=500.
Use set.seed(11) Start with simulation of u.
For each simulated value u generate v from the corresponding conditional distribution p(v|u)).
Finally, for each pair v,u simulate w from p(w|v,u).
```{r}
set.seed(11)
sim.u <- rbinom(500, size=1, p=.45)
sim.v<-sapply(sim.u,function(z) sample(1:3,1,replace=T,
                                       prob=ifelse(rep(z==1,3),pCond.v.given.u1,pCond.v.given.u0)))
sim.w<-apply(cbind(sim.u,sim.v),1,
              function(z) sample(1:3,1,replace=T,prob=switch(3*z[1]+z[2],p.given.u0.v1,p.given.u0.v2,
                                                             p.given.u0.v3,p.given.u1.v1,p.given.u1.v2,
                                                             p.given.u1.v3)))
```

# 3 Conditional Expected Values

Calculate unconditional expected value E[v]
Random variable v can take values 1,2,3 with corresponding probabilities
```{r}
vMarginal
```
Then the unconditional mean value is
```{r}
c(1,2,3)%*%vMarginal
```
Calculate conditional expected Value E[v|u]

First, find conditional mean values E[v|u=u0]=E[v|u0] and E[v|u=u1]=E[v|u1].

The random variable (v|u0)takes values 1,2,3with corresponding probabilities p(v=1|u0),p(v=2|u0),p(v=3|u0), given by the vector
```{r}
(cond.v.given.u0<-apply(data3way.array.p[,,"u0"],2,sum)/uMarginal["u0"])
```
Taking conditional expected value with respect to this distribution E[v|u0]E[v|u0] is:
```{r}
(exp.v.given.u0<-c(1,2,3)%*%cond.v.given.u0)
```


The random variable (v|u1) takes the same values 1,2,3, but with different probabilities p(v|u1):
```{r}
cond.v.given.u1
```
Thus conditional expected value E[v|u1] is
```{r}
(exp.v.given.u1<-c(1,2,3)%*%cond.v.given.u1)
```

Note that the conditional expected value E[v|u] takes two different values: E[v|u0] and E[v|u1] with probabilities p(v|u0) and p(v|u1), correspondingly.

This means that E[v|u] is a random variable and its (unconditional) expected value is
E[v|u]=E[v|u0]p(u=u0)+E[v|u1]p(u=u1).
Calculate this unconditional expected value:
```{r}
(uncond.exp.v.given.u<-exp.v.given.u0*uMarginal["u0"]+exp.v.given.u1*uMarginal["u1"])

```
